---
title       : Session7
subtitle    : 
author      : 
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

<style>
strong {
  font-weight: bold;
  color: red;
  font-size: 115%
}
</style>



--- .segue bg:grey

# Exkurs

Graphics (line)



--- .segue bg:grey

# Recap


---

## Recap


You should know
  - what factor means in R
  - how to create factors


---

## Recap



```{r}
library(readxl)
babies <- read_excel("../session7dta/babies.xlsx",1)
babies$sex.mf <- factor(babies$sex, 
                        levels = c(0,1),
                        labels = c("male","female"))
```


--- 

## Recap factors 

- use `factor()` to change the `marital` and `inc` variable from numeric to factor
  - 1 = married, 2 = legally separated, 3 = divorced, 4 = widowed, 5 = never married
  - 0 = <2500, 1 = 2500-4999, 2 = 5000-7499, 3 = 7500-9999, ..., 9 = 15000+, 98 = unknown, 99 = not asked
- table the two variables using `table()` and `prop.table()`; what is the percentage of <2500 amongst married and what amongst never married?


---


## Solution

```{r}
babies$marital <- factor(babies$marital, 
                         levels = 1:5, 
                         labels = c("married","legally separated",
                                    "divorced","widowed","never married"))
babies$inc <- factor(babies$inc,
                     levels = c(0:9,98,99),
                     labels = c("<2500",
                                paste(seq(2500,20000,by = 2500),
                                       seq(4999,22500,by = 2500),sep = "-"),
                                "25000+","unknown","not asked"))
```



---


## Solution
```{r}
addmargins(prop.table(table(babies$inc,babies$marital),2))
```



---


## change labels

```{r,echo=FALSE}
babies$sex.mf <- factor(babies$sex, 
                        levels = c(0,1),
                        labels = c("male","female"))
```


```{r}
table(babies$sex.mf)
babies$sex.mj <- factor(babies$sex.mf, 
                        levels = c("female","male"),
                          labels = c("Maedchen","Junge"))
table(babies$sex.mj,babies$sex.mf)
```



---


## Exercise

Recode the sex variable again. Use `boy` and `girl` as label.


---


## recoding

>  - another kind of issue is a problem like the following:
>  - the `drace` variable contains 11 races plus one unknown coding
>  - coding 0-5 means all white, 6 mex, 7 black, 8 asian, 9 and 10 mixed, 99 unknown



---


```{r}

library(car)
babies$drace <- recode(babies$drace,
                       '0:5="white";6="mex";7="black";
                       8="asian";c(9,10)="mixed";99=NA')
```



---


## Exercise

- recode `drace` in the same way
- use the variables `race` and `drace` to get the percentage of mixed paires (man and woman from different races)


---


## Solution

```{r}
prop.table(table(babies$race == babies$drace))
```


---


## recoding missings

- remember how we used numbers to extract parts of vectors and data frames, e.g.


```{r}
babies$id[1] ## the first element of the id-row
```



```{r}
babies[2,1] ## the second element of the first column
```


---


```{r}
babies[1:2,] ## the first two rows
```


---


```{r}
babies[1:2,c(1,3)] ## the first and third column for the first two rows
```


---


## Logical Indexing

>  - Problem: get all ids of babies with a birth weight > 4
>  - Solution: instead of numbers as indices we use conditions 



---


## Logical Indexing

- Problem: get all ids of babies with a birth weight > 4
- Solution: instead of numbers as indices we use conditions 

```{r}
babies$id[babies$bwt > 4]
```


---


## Logical Indexing

- Problem: get all rows of babies with a birth weight > 5

```{r}
babies[babies$bwt > 5,]
```


---


## Logical Indexing

- Problem: get all rows of babie girls with a birth weight > 4.5

```{r}
babies[babies$bwt > 4.5 & babies$sex.mf == "female",]
```



---


## Recode Missings

Problem: the `age` and the `dage` column contain 99 to indicate missing. In R a missing is indicated by `NA`.  


```{r}
summary(babies$age)
```


--- 


## Recode Missings

Problem: the `age` and the `dage` column contain 99 to indicate missing. In R a missing is indicated by `NA`.  


```{r}
babies$age[babies$age == 99]
```


--- 


## Recode Missings

Problem: the `age` and the `dage` column contain 99 to indicate missing. In R a missing is indicated by `NA`.  


```{r}
babies$age[babies$age == 99] <- NA
```



```{r}
summary(babies$age)
```



---

## Exercise


Recode the 99 by `NA`s in the `dage` column of the data set.



---

## Solution


Recode the 99 by `NA`s in the `dage` column of the data set.


```{r}
babies$dage[babies$dage == 99] <- NA
```


```{r}
summary(babies$dage)
```


---

## cut a numeric vector


Another way of recoding often used is the transformation of a numeric vector into a factor by cutting the numeric vector at specific cut points (breaks).

For example we want to transform the fathers age variable into five-years intervals:

```{r}
summary(babies$age)
```


```{r}
babies$dage.cat <- cut(babies$dage,
                      breaks = seq(15,45,by = 5), 
                      include.lowest = T)
```



---

## Exercise

Cut also the mother's age into five-years intervals. Make a boxplot with mother's age (use the age groups) on the x-axis and 
birth weight on the y-axis.




